---
title: "Barcoder Parameters"
subtitle: "Created July 17, 2020"
author: "PME"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding=encoding, output_dir=here::here('Results'))})
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: hide
  pdf_document: default
  word_document: default
---

```{r set_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, 
                      warning=FALSE,
					  fig.align='center')
```

guppy_barcoder has a number of options to optimize binning. These include:

- `--min_score` (60)
- `--require_both_ends` (FALSE)
- `--min_score_rear_override`
- `--detect_mid_strand_barcodes` (FALSE)
- `--min_score_mid_barcodes` (50)
- `--front_window_size` (50)
- `--rear_window_size`

Gap penalties can also be modified for starting (40), opening (40), extending (40), and ending (40). 

The scores are calculated by modified Needleman-Wunsch algorithm, and normalized by the length of the barcode sequence. Scores for matches and mismatches are (available in /opt/ont/guppy/data/barcoding): 

`
   	  A	     C	   G	   T	N
  A	 96	  -316	-192	-369	0
  C	-316	 100	-352	-295	0
  G	-192	-352  	98	-329	0
  T	-369	-295	-329	 100	0
  N	   0	   0	   0	   0	0
`

Because scores are normalized, the highest possible score is 100. The highest score barcode is considered a candidate match, which if above `--min_score`, is considered successful. 

Guppy_barcoder produces a summary table for each read, which includes the read id, barcode matches, sequences, and scores, etc. We'll use that to set a proper threshold given our knowledge of the samples. 

```{r}
libs = c('lattice', 'here', 'magrittr')

for (i in libs) library(i, character.only=TRUE)

df = here('Data', 'barcoding_summary.txt') %>% 
  read.table(header=TRUE)
```

